Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-14 sports_analytics.Rmd
added logos to Figure 2; #10

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
  - name: Katherine Evans
    affiliation: Monumental Basketball
    location: Washington, DC 20003
    email: kevans@monumentalsports.com
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data. Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights. This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: rticles::arxiv_article
header-includes:
  - \usepackage{amsmath}
---

@@

@@

## Graphical/Visual Abstract and Caption

> Include an attractive full color image to go under the text abstract and in the online Table of Contents.
You will also need to upload this as a separate file during submission.
It may be a figure or panel from the article or may be specifically designed as a visual summary. While original images are preferred, if you need to look for a thematically appropriate stock image, you can go to pixabay.com (not affiliated with Wiley) to find a free stock image with a CC0 license. Another option you have is to utilize professional illustrators with Wiley’s Graphical Abstract Design service.

> Size: The minimum resolution is 300 dpi. Please keep the image as simple as possible because it will be displayed in multiple sizes. Multiple panels and text other than labels are strongly discouraged.

> Caption: This is a narrative sentence to convey the article's essence and wider implications to a non-specialist audience. The maximum length is 50 words, but consider using 280 characters or less to facilitate social media sharing, which can increase the discoverability of your article.

@@

vikings <- teamcolors |> 
  filter(league == "nfl", name == "Minnesota Vikings") |> 
  pull(primary)

saints <- teamcolors |> 
  filter(league == "nfl", name == "New Orleans Saints") |> 
  pull(primary)
```

```{r nflfastr_wp, fig.cap="Win probability graph for Saints vs. Vikings in the 2017--18 NFL Playoffs."}
library(nflfastR)
x <- load_pbp(2017) |> 
  filter(
    season_type == "POST",
    home_team == "MIN",
    away_team == "NO",
    !is.na(home_wp),
    !is.na(away_wp)
  )
x |>
  select(game_seconds_remaining, home_wp, away_wp) |> 
  pivot_longer(
    !game_seconds_remaining,
    names_to = "team",
    values_to = "wp"
  ) |>
  ggplot(aes(game_seconds_remaining, wp, color = team)) +
  scale_x_reverse(breaks = seq(0, 3600, 900)) + 
  scale_color_manual(
    values = c(saints, vikings), 
    labels = c("New Orleans Saints", "Minnesota Vikings")
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(size = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

@@

@R-teamcolors provides color palettes for professional and amateur sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
Figure \ref{fig:teamcolors} illustrates team color palettes for professional sports leagues.

>>>>

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
  - name: Katherine Evans
    affiliation: Monumental Basketball
    location: Washington, DC 20003
    email: kevans@monumentalsports.com
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. We connect four(?) big ideas that are common across multiple sports, and explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves both statistical methodologies and sources of data. Where appropriate, we outline how data, models, and knowledge of the sport combine to generate actionable insights. This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: rticles::arxiv_article
header-includes:
  - \usepackage{amsmath}
---

@@

<!--

@@

-->

@@

miracle <- teamcolors |> 
  filter(league == "nfl", grepl("Vikings|Saints", name)) |>
  arrange()
```

```{r nflfastr_wp, fig.cap="Win probability graph for New Orleans Saints vs. Minnesota Vikings in the 2017--18 NFL Playoffs."}
library(nflfastR)
x <- load_pbp(2017) |> 
  filter(
    season_type == "POST",
    home_team == "MIN",
    away_team == "NO",
    !is.na(home_wp),
    !is.na(away_wp)
  ) |>
  select(game_seconds_remaining, home_wp, away_wp) |> 
  pivot_longer(
    !game_seconds_remaining,
    names_to = "team",
    values_to = "wp"
  )

wp_plot <- ggplot(x, aes(game_seconds_remaining, wp, color = team)) +
  scale_color_manual(
    values = setNames(miracle$primary, c("home_wp", "away_wp")), 
    labels = miracle$name
  ) +
  geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") +
  geom_vline(
    xintercept = seq(0, 3600, 900), 
    linetype = "dashed", 
    color = "gray"
  ) + 
  geom_line(size = 1.5, alpha = 0.8) +
  labs(
    x = "Time Remaining (seconds)",
    y = "Win Probability",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")

library(magick)
logos <- miracle |>
  pull(logo) |>
  map(image_read) |>
  map(as.raster)

wp_plot + 
  annotation_raster(logos[[1]], -1010, -1490, 0.76, 0.865) + 
  annotation_raster(logos[[2]], -1010, -1490, 0.135, 0.24) + 
  scale_x_reverse()
```

@@

@R-teamcolors provides color palettes for professional and amateur sports teams, as well as color and fill scale functions compatible with \pkg{ggplot2} [@R-ggplot2].
The team colors and logos shown in Figure \ref{fig:nflfastr_wp} were provided by the \pkg{teamcolors} package. 
Figure \ref{fig:teamcolors} illustrates team color palettes for professional sports leagues.
